Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-13 SCP2008/paper/metamonads.tex
many comments to Section 2. I believe it needs abbreviation, to keep it more technical.

\documentclass[draft]{elsart}

@@

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
% \setlength{\parskip}{0pt}
\newlength{\ourmathindent}
\setlength{\ourmathindent}{1em}

@@

\begin{abstract}
We use multi-stage programming, monads and OCaml's
advanced module system to demonstrate how to eliminate all
abstraction overhead while avoiding any inspection of the resulting
code.  We demonstrate this clearly with LU decomposition as a 
representative family of symbolic and numeric algorithms. 
We parameterize our code to a great extent
(over domain, matrix representations, determinant tracking, 
pivoting policies, result types, etc) at no run-time cost.  Because
the resulting code is generated just right and not changed afterwards,
MetaOCaml guarantees that the generated code is well-typed.
We further demonstrate that various abstraction parameters (aspects)
can be made orthogonal and compositional, even in the presence of
name-generation for temporaries, and 
``interleaving'' of aspects.  We also show how to encode some
domain-specific knowledge so that ``clearly wrong'' compositions can
be rejected at or before generation time, rather than during
the compilation or running of the generated code.
\end{abstract}

@@

In high-performance symbolic and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. Implementations of the
widely used LU decomposition algorithm (which subsumes Gaussian 
Elimination (GE)) --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the determinant or rank, how
and whether search for a pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\footnote{We describe here an improved version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have also changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of LU decomposition 
and linear system solving as our running examples to demonstrate our
techniques.  Specifically, our contributions:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,MSP:PADL04} for generating control structures
      such as loops and conditionals. The extension is non-trivial
	  because of control dependencies and because let-insertion, as we argue,
	  is a control effect on its own: for example\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

We also used the same technology to implement a Runge-Kutta solver for
ordinary differential equations, as well as a reimplementation of
the FFT algorithm from~\cite{KiselyovTaha}.  The technology presented
here was amply sufficient for these implementations.  Since our current
implementations of these algorithms are rather straightforward compared to 
our versions of LU Decomposition, these will not be mentioned further
in this paper.

@@

Before studying the details of the technologies involved in the 
implementation, it is worthwhile to carefully study the design space involved.
A preliminary study~\cite{Carette06} revealed a number of aspects of the 
family of Gaussian Elimination algorithms.  In the present work, we outline
a number of additional aspects involved in the faimily of LU decomposition 
algorithms.  These will first be presented in a somewhat ad hoc manner, 
roughly coresponding to the order in which they were ``discovered''.  These
will then be reorganized into sets of semantically related aspects, which will
form the basis of our design.  Finally, we will extract a methodology from
our experience.

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  When pivoting is used, we get a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition, with $L$ remaining unit lower-triangular but with a
$U$ which is no longer upper triangular but instead ``staggered''
in the upper triangle.

@@

We reuse the english word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While this use if very much in the same 
spirit as in aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
our implementation methodology is radically different\footnote{However it seems
that we are closer to the original ideas on AOP~\cite{709568,mendhekar97rg}
which were also concerned with scientific software}.  We firmly believe that
our typed generative methodogy is much better suited to the functional
programming paradigm than attempting to graft the program-trace-based
methodology of object-oriented versions of AOP.

@@

		was represented as an array of arrays, a one-dimensional array,
		a hash table, etc.  For the case of a one-dimensional array,
		whether indexing was done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this was sometimes used.

@@

		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this was sometimes used to choose

@@

	\item \textbf{Output choices}:  Whether just the reduced matrix
		(the `U' factor), both L and U, as well as
		the rank, the determinant, and/or the sequence of 
		pivots is to be returned.  For example, Maple's

@@

	\item \textbf{Code representation}: Whether we are actually generating
		code or we are in fact representation the algorithm in ``direct
		style'', albeit with a lot of abstraction overhead.

@@

		routine needs to be used.  It turns out that for certain classes
		of \emph{expressions}, it is convenient to use a zero-equivalence
		test which is separate from the domain normalization.  This is

@@

	  as a list of row (and/or column) exchanges, as a unitary matrix,

@@

@@

There are other aspects which we have not (yet) implemented, but that have
been seen in different implementations.  In particular,
\begin{enumerate}
\setcounter{enumi}{\value{naspects}}
\item \textbf{Logging}: This is a classical cross-cutting concern, which
is also useful in this context.  Of course, there are many different
parts of the algorithm which could be ``logged'', and ensuring fine control
over this is a thorny issue.
\item \textbf{Sparsity}: If a matrix is known to be sparse, the traversal
(at least) should be sparse.  Additionally, it would better still if
care was taken to minimize fill-in.
\item \textbf{Other structure}: For example, if a matrix is known in
advance to be real symmetric tri-diagonal, then LU decomposition can be
done in $O(n^2)$ time (instead of $O(n^3)$; the main difficulty in 
implementing this kind of reduction is the fact that this version of 
LU requires $O(n)$ extra storage (unlike all other versions).
\item \textbf{Warnings}: In the case of LU decomposition of a matrix over
a domain for which we only have a heuristic algorithm for zero
testing, it is customary to issue a warning (or otherwise log) when 
a potentially zero pivot is chosen.
\item \textbf{In-place}: When it is possible, it would be good to have the
option to re-use the input matrix as the storage space for the output.
\item \textbf{Error-on-singular}: There are instances where we are only
interested in an LU decomposition for a non-singular matrix, and would
prefer to get an exception when a singular matrix is encountered.
\setcounter{naspects}{\value{enumi}}
\end {enumerate}

Clearly the above $\thenaspects$ aspects have a rather complex 
dependency graph.  While some aspects are purely orthogonal to others, 
most are inter-dependent.  For example, if one wants the determinant of
the matrix as part of the output, then the determinant has to be tracked
during the algorithm; a priori, it is possible to make incompatible choices,
and our implementation has to insure that this is in fact impossible.
More precisely, our goal is to make sure that this is detected long
before the algorithm could be run.

There are other kinds of dependencies: for example, whether or not a 
domain has a length measure will influence pivoting, \emph{if} pivoting
is to be performed.  While there are no circular dependencies, there is
no clear directed graph of dependencies.  In other words, one cannot 
simply use product, co-products, and compositions of the various aspects 
to build the resulting algorithm.  We will get back to this point when
we describe the details of the implementation.

@@

Further investigation of the aspects reveals that these fall into
various categories.  This categorization has already been useful to us,
as it revealed that parts of our implementation was not as well
separated into distinct concerns as we had originally thought.  In
particular, it revealed that our implementation of ``containers'' 
mixed representation (i.e. data-structure) issues with abstraction
issues (i.e. matrices and their natural representation-independent
operations).  We hope to rectify this in the future.

\begin{enumerate}
	\item \textbf{Abstract domain}.  In other words, these are
		firmly aspects which are domain-specific, and encode
		requirements level entities.  In particular, items like
		Domain, length measure, Matrix, permutation, full division
		and structured matrices all belong here.
	\item \textbf{Concrete representation}.  In other words, the
		design-time issue of choosing a data-structure to represent
		various abstract domains.  Items like whether the results 
		should be packed, (part of) the representation of containers,
		sparsity, and representation of permutations.
	\item \textbf{Interface}.  In other words, what will be the
		interface of each generated function.  This mostly deals
		with the input and output choices, but also includes whether
		the output should be in-place, and whether errors or warnings
		should be issued in certain situations.  Logging is also
		an interface issue.
	\item \textbf{Algorithmic Strategy}.  This encodes the various
		choices in the algorithms themselves.  For example, whether
		to use fraction-free updates, which pivot strategy to use,
		whether the matrix is augmented.
	\item \textbf{Information tracking}.  What information should be
		kept and/or computed while the algorithm proceeds, such
		as whether to track determinant, rank, or pivot.
	\item \textbf{Staging}.  This is the abstract interface to the
		underlying programming language.  It (currently) encodes
		whether to generate code that will be executed in a future
		stage, or whether the code is direct and should be executed
		now.  It should be possible to encode other strategies too.
\end{enumerate}

While the first two families of aspects are the most obvious, it is quite
difficult to separate them out cleanly.  This seems to stem from the fact
that in a non-staged setting, separating out these aspects leads to such
fantastically inefficient code, especially in the setting of scientific
computation, that our experience works \emph{against us}.  While it would
be too much to speak of a paradigm shift, it is necessary to firmly think
in terms of abstract ideal design while avoiding all thoughts of traditional
efficiency.

Interestingly, unlike traditional approaches \cite{journals/cacm/parnas72a}, we
allow the interfaces to our (generated) routines to vary.  Compared to dynamic
approaches in Object-Oriented language (late binding and dynamic dispatch) or
functional languages (Haskell's dictionary-based \emph{type classes}), which
offer some flexibility on interfaces, our interfaces are actually fixed at
run-time, thus do incur any run-time inefficiency whatsoever.

While we are by no means the first to speak of algorithmic strategy, to
our knowledge our approach is one of the cleanest, in that the issue of
strategy is completely separated into an independent entity (module), and
yet we suffer no run-time inefficiency because of it.

The information tracking aspects are in some ways to most intruiging ones:
they are purely internal to the algorithms, and they are the ones which
are most dependent on other choices.  The \emph{implementation} of these
aspects is completely orthogonal to the other aspects, however their
\emph{compositionality properties} are deeply enmeshed with that of other
aspects.  While writing the code for the aspects was straightforward enough,
this is where quite a bit of infrastructure was needed to encode
various dependencies, so that it would be impossible to generate an
instance of the LU Decomposition algorithm with information
tracking aspects which were incompatible with the requirements of other
aspects.

Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
we abstracted out the underlying programming language completely.  This 
allowed us to both generate (efficient) code and to write a directly
runnable (highly inefficient) version of the algorithm from the same
generator.  This brought out the issue that, once this was done, we had
essentially abstracted out all of the syntactic sugar of the underlying
programming language, and all we were left with was function application
and monadic composition.  This means the code for our generator looks mostly
like Scheme with added syntax for monads!

@@

Our overall approach can best be described a combination 
of two approaches: hand-writing a code generator (cogen) suitable
for multi-level specialization \cite{Gluck95,GluckJ97} and creating
an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
More specifically we start from a \emph{known} set of implementations
of an algorithm, and extract commonalities and variation points.
This is unlike~\cite{journals/tse/Parnas76} and most subsequent
approaches to product families, as we do not over-engineer our design
by imagining weird variations that are unlikely to come up in realistic
situations, but only create variations when we notice them in actual use.
This approach is quite well-suited to the development of scientific
software, which has a rich history and where most useful variations have
already appeared in some form.

>>>>

\documentclass[draft]{elsart}
\hyphenation{ana-lysis theo-ry}

@@

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
% \setlength{\parskip}{0pt}
\parskip=0pt
\newlength{\ourmathindent}
\setlength{\ourmathindent}{1em}

@@

\begin{abstract}
We use multi-stage programming, monads and OCaml's
advanced module system to demonstrate how to eliminate all
abstraction overhead while avoiding any inspection of the resulting
code.  We demonstrate this clearly with LU decomposition as a 
representative family of symbolic and numeric algorithms. 
We parameterize our code to a great extent --
over domain, input and permutation matrix representations, 
determinant and rank tracking, 
pivoting policies, result types, etc -- at no run-time cost.  Because
the resulting code is generated just right and not changed afterwards,
MetaOCaml guarantees that the generated code is well-typed.
We further demonstrate that various abstraction parameters (aspects)
can be made orthogonal and compositional, even in the presence of
name-generation for temporaries, and 
``interleaving'' of aspects.  We also show how to encode some
domain-specific knowledge so that ``clearly wrong'' compositions can
be rejected at or before generation time, rather than during
the compilation or running of the generated code.
\end{abstract}

@@

In high-performance symbolic and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. Implementations of the
widely used LU decomposition algorithm and its particular case Gaussian 
Elimination (GE) \oleg{can I say GE is an instance of LU?} --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the determinant or rank, how
and whether search for a pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\footnote{We describe here an new version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have also changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of LU decomposition 
and linear system solving as our running examples to demonstrate our
techniques.  Specifically, our contributions:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,MSP:PADL04} for generating control structures
      such as loops and conditionals. The extension is non-trivial
	  because of control dependencies and because let-insertion, as we argue,
	  is a control effect on its own: for example
      \texttt{let x = exp in \dots} has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

We also used the same technology to implement a Runge-Kutta solver for
ordinary differential equations, as well as a reimplementation of
the FFT algorithm from~\cite{KiselyovTaha}.  The technology presented
here was amply sufficient for these implementations.  Since our current
implementations of these algorithms are rather straightforward compared to 
our versions of LU Decomposition, these will not be mentioned further
in this paper (the code is available at \cite{metamonadsURL}).

@@

Before investigating implementation approaches, it is worthwhile to
carefully study the design space involved.  A preliminary
study~\cite{Carette06} revealed a number of aspects of the family of
Gaussian Elimination algorithms.  In the present work, we outline a
number of additional aspects involved in the faimily of LU
decomposition algorithms.  These will first be presented in a somewhat
ad hoc manner, roughly coresponding to the order in which they were
``discovered''.  We then reorganize them into groups of
semantically related aspects to form the basis of our design.
Finally, we will extract a methodology from our experience.

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  When pivoting is used, we get a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
$U$ is no longer upper triangular but rather ``staggered''
in the upper triangle.

@@

We reuse the English word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While our use shares the 
spirit of aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
our implementation methodology is radically different\footnote{However it seems
that we are closer to the original ideas of AOP~\cite{709568,mendhekar97rg}
which were also concerned with scientific software}.  We firmly believe that
our typed generative methodogy is much better suited to functional
programming, compared to attempts to graft the program-trace-based
methodology of object-oriented versions of AOP.

@@

		was represented as an array of arrays, a
		one-dimensional array with C or Fortran indexing styles, 
		a hash table, etc. Efficient row exchanges, if available for a
		particular representation, were sometimes used.

@@

		(whether numerical or coefficient growth), if a domain 
	  possesses
		an appropriate length measure, it was sometimes used to choose

@@

	\item \textbf{Output choices}:  Just the reduced matrix
		(the `U' factor) or both L and U factors. The output
	  choices also include
		the rank, the determinant, and the sequence of 
		pivots.  For example, Maple's

@@

	\item \textbf{Code representation}: the form and the language
	  for the generated code (OCaml, C, Fortran, etc). A
	  degenerate case, useful for testing, is for the generator
	  to run the LU algorithm directly (albeit with great
	  abstraction overhead).

@@

		routine.  For certain classes
		of \emph{expressions}, it turns out convenient to use 
		a zero-equivalence
		test that is separate from the domain normalization.  This is

@@

	  as a list of row and column exchanges, as a unitary matrix,

@@

	  \oleg{Is this the same as Augmented matrices above?}

@@

The aspects in the following group have also been observed in
practice but we have not yet implemented them:
\begin{enumerate}
\setcounter{enumi}{\value{naspects}}
\item \textbf{Logging}, a classical cross-cutting concern.
\item \textbf{Sparsity}: If a matrix is known to be sparse, at least
the traversal should be sparse. Maximal preservation of the sparsity
is desirable. 
\item \textbf{Other structure}: If a matrix is known in
advance to be real symmetric tri-diagonal, LU decomposition can be
done in $O(n^2)$ rather than $O(n^3)$ time, at the additional $O(n)$
storage cost.
\item \textbf{Warnings}: If a domain permits only heuristic zero
testing, it is customary to issue a warning (or otherwise log) when 
a potentially zero pivot is chosen.
\item \textbf{In-place}: offering an option of in-place decomposition,
  re-using the input matrix as the storage for the output.
\item \textbf{Error-on-singular}: Raise an exception when the input
  matrix is (near) singular.
\setcounter{naspects}{\value{enumi}}
\end {enumerate}

Most of these aspects are inter-dependent. For example, if the
determinant is the part of the output, the determinant should be
tracked during the decomposition. Determinant should also be tracked
if the fraction-free aspect is chosen. The availability of the length
measure in the domain influences pivoting, \emph{if} pivoting
is to be performed. One could therefore select aspects that turn out
incompatible, and we are to prevent this. 
More precisely, our goal is to detect the selection of
incompatible aspects long before the generated code is run.

\omitnow{Clearly the above $\thenaspects$ aspects have a rather complex 
dependency graph.  While some aspects are purely orthogonal to others, 
most are inter-dependent.  For example, if one wants the determinant of
the matrix as part of the output, then the determinant has to be tracked
during the algorithm; a priori, it is possible to make incompatible choices,
and our implementation has to insure that this is in fact impossible.
There are other kinds of dependencies: for example, 
  While there are no circular dependencies, there is
no clear directed graph of dependencies.  In other words, one cannot 
simply use product, co-products, and compositions of the various aspects 
to build the resulting algorithm.  We will get back to this point when
we describe the details of the implementation.}

@@

\oleg{I'm quite uneasy about this and the following sections. It
  appears as `philosophy'. It's OK for us, because it is written in
  the Perfect tense for us. We understand what each of these items
  mean. OTH, readers will read this in the Future tense. They have not
  seen our code and know nothing about our approach. To them, it may
  appear just opaque philosophising.

I'm in favor of drastically abbreviating these sections. Our message
to the reader should be: please keep these groups of aspects as you
read the description of our code below. We also point out that
attempting to separate these aspects and letting the users
independently change them leads -- in the unstaged case -- to such
fantastically inefficient code so that such clean design is rarely
(cite Aldor?) attempted. Staging let us remove the performance
penalties of abstraction and so let us concentrate on the design. 
The last phrase is actually more suitable for conclusion or related
work. I think it would be much better to let related work make these
points for us.}

Further investigation revealed the following grouping of aspects:%
\footnote{This grouping showed that some of our implementation did not
separate distinct concerns well. In
particular, our implementation of ``containers'' 
mixed representation (i.e. data-structure) issues with abstraction
domain issues.  We hope to rectify this in the future.}

\begin{enumerate}
	\item \textbf{Abstract domain}.  This group includes
	  `mathematical' aspects: domain of matrix elements, length
	  measure, full division, normalization, zero testing,
	  symmetry and other matrix structure.
	\item \textbf{Concrete representation}: choosing data
	  structures for domains, containers, permutation
	  matrices. This group also includes packing, in-place
	  decomposition, and the
	  representation of sparse matrices.
	\item \textbf{Interface} of each generated function, including
	  the input and output choices, logging and error reporting.
		\oleg{In-place updates is also the issue of concrete
		  representation! As well as of the abstract domains:
		  for some domains we just can't run LU with no extra
		  storage.} 
	\item \textbf{Algorithmic Strategy}, such as fraction-free
	  updates, pivoting strategies, augmented matrices.
	\item \textbf{Tracking}, of determinant, rank, or pivot, etc.
	\item \textbf{Staging}: whether the result if the program or a
	  generator that will produce the program.
\end{enumerate}

\oleg{I'd rather replace this paragraph with my remark above.}
While the first two families of aspects are the most obvious, it is quite
difficult to separate them out cleanly.  This seems to stem from the fact
that in a non-staged setting, separating out these aspects leads to such
fantastically inefficient code, especially in the setting of scientific
computation, that our experience works \emph{against us}.  While it would
be too much to speak of a paradigm shift, it is necessary to firmly think
in terms of abstract ideal design while avoiding all thoughts of traditional
efficiency.

\oleg{I'd rather move this down where we discuss the output choices. I
  want to get top the code quickly. Actually, the following paragraph
  better belongs to related work.}  
Unlike traditional approaches
\cite{journals/cacm/parnas72a}, the interfaces of our generated
routines vary depending on the choices of input and output
aspects. Dynamic approaches in Object-Oriented language (late binding
and dynamic dispatch) or functional languages (Haskell's
dictionary-based \emph{type classes}), too, offer flexibility of
interfaces. In our approach, however, it is the generator that
produces code whose interfaces depend on the arguments of the
generator. The interfaces of the generated routines are all fixed and
hence effecient.

\oleg{I'd drop this: the reader has not seen our implementation.}
The information tracking aspects are in some ways to most intruiging ones:
they are purely internal to the algorithms, and they are the ones which
are most dependent on other choices.  The \emph{implementation} of these
aspects is completely orthogonal to the other aspects, however their
\emph{compositionality properties} are deeply enmeshed with that of other
aspects.  While writing the code for the aspects was straightforward enough,
this is where quite a bit of infrastructure was needed to encode
various dependencies, so that it would be impossible to generate an
instance of the LU Decomposition algorithm with information
tracking aspects which were incompatible with the requirements of other
aspects.

\oleg{I'd move this to the next section.}
Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
we abstracted out the underlying programming language completely.  This 
allowed us to both generate (efficient) code and to write a directly
runnable (highly inefficient) version of the algorithm from the same
generator.  This brought out the issue that, once this was done, we had
essentially abstracted out all of the syntactic sugar of the underlying
programming language, and all we were left with was function application
and monadic composition.  This means the code for our generator looks mostly
like Scheme with added syntax for monads!

@@

Our overall approach can best be described a combination 
of two approaches: hand-writing a code generator (cogen) suitable
for multi-level specialization \cite{Gluck95,GluckJ97} and creating
an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
\oleg{I'd keep only the above phrase and say that our approach of
  using staging, monads and functors turns out to permit extensible
  set of aspects and flexible enough for iterative improvement in the
  design.}

\oleg{I'm inclined then to drop the following two paragraphs. I'm
  positive reviewers would be upset at philosophical poinst in the
  middle of a technical paper.}
More specifically we start from a \emph{known} set of implementations
of an algorithm, and extract commonalities and variation points.
This is unlike~\cite{journals/tse/Parnas76} and most subsequent
approaches to product families, as we do not over-engineer our design
by imagining weird variations that are unlikely to come up in realistic
situations, but only create variations when we notice them in actual use.
This approach is quite well-suited to the development of scientific
software, which has a rich history and where most useful variations have
already appeared in some form.
